Individual and Differential Harm in Redistricting

### 2022-12-06 paper/harm.Rmd
str_replace electoral system

For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [@griffith1907rise, pp. 72--73], is perhaps the most consistently visible manifestation of this tendency. But gerrymandering takes other forms as well. In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial]. At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process]. The diversity of types of gerrymandering is further apparent from a survey of major litigation over the past decades, which includes allegations of gerrymandering involving party [@2018league; @2019rucho], race [@1993shaw; @2022merrill], religion [@1977ujo; @2020ercsd], incarceration status [@2015calvin; @2016davidson], and residency or citizenship [@1966burns; @2016evenwel].

The academic study of gerrymandering has unfortunately not reflected this diversity.
The vast majority of research attention has been on partisan gerrymandering, and has particularly centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020]. A large number of formulas for quantifying or identifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18]. The study of racial gerrymandering has been largely shaped by the U.S. legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA). This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological]. Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].

@@

Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
Many factors influence the performance and qualities of an electoral system.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

The combination of individual voter choices and counterfactual electoral systems defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
Section \ref{sec:motivation} illustrates this with a simple example, and the general definitions are formalized in Section \ref{sec:indiv-harm}.

@@

These implicit counterfactuals differ from the explicit counterfactuals we adopt in this work in that there is no guarantee they are possible.
@katz2020 and others recognize that a nonzero value in a measure like partisan bias can only be considered problematic "if there is a fairer alternative." The inability of partisan bias and other existing measures to explicitly account for realistic counterfactual districting plans means that none of them may be interpreted in a vacuum.
No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the district-based electoral system.

@@

At its core, redistricting is a multifaceted and high-dimensional problem that must balance political, legal, and administrative constraints to allow citizens to exercise their rights.
A districting plan can never be boiled down to a single number.
But we hope our approach provides a modular and flexible set of measures that can help decompose how various components of the electoral system combine to produce observed patterns and election outcomes.
Researchers can and should tinker with our assumptions and push our framework further to better apply it to multi-candidate elections, more realistic election models, and beyond district-based electoral systems.

>>>>

For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [@griffith1907rise, pp. 72--73], is perhaps the most consistently visible manifestation of this tendency. But gerrymandering takes other forms as well. In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial]. At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the districting plan may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process]. The diversity of types of gerrymandering is further apparent from a survey of major litigation over the past decades, which includes allegations of gerrymandering involving party [@2018league; @2019rucho], race [@1993shaw; @2022merrill], religion [@1977ujo; @2020ercsd], incarceration status [@2015calvin; @2016davidson], and residency or citizenship [@1966burns; @2016evenwel].

The academic study of gerrymandering has unfortunately not reflected this diversity.
The vast majority of research attention has been on partisan gerrymandering, and has particularly centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020]. A large number of formulas for quantifying or identifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18]. The study of racial gerrymandering has been largely shaped by the U.S. legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA). This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological]. Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].

@@

Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
Many factors influence the performance and qualities of a districting plan.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

The combination of individual voter choices and counterfactual districting plans defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
Section \ref{sec:motivation} illustrates this with a simple example, and the general definitions are formalized in Section \ref{sec:indiv-harm}.

@@

These implicit counterfactuals differ from the explicit counterfactuals we adopt in this work in that there is no guarantee they are possible.
@katz2020 and others recognize that a nonzero value in a measure like partisan bias can only be considered problematic "if there is a fairer alternative." The inability of partisan bias and other existing measures to explicitly account for realistic counterfactual districting plans means that none of them may be interpreted in a vacuum.
No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the electoral system's geography.

@@

At its core, redistricting is a multifaceted and high-dimensional problem that must balance political, legal, and administrative constraints to allow citizens to exercise their rights.
A districting plan can never be boiled down to a single number.
But we hope our approach provides a modular and flexible set of measures that can help decompose how various components of the districting plan combine to produce observed patterns and election outcomes.
Researchers can and should tinker with our assumptions and push our framework further to better apply it to multi-candidate elections, more realistic election models, and beyond district-based electoral systems.
